One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-08-03 papers/1_dmatpol/paper-submit.tex
Submition version added

The electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix 
susceptibilities (DMS) interacting with the spatial distribution of electric field.
The analytical approximate expressions are derived at the Hartree\hyp{}Fock theory, 
which serves as a basis for construction of generalized model 
that can be used for an arbitrary form of wavefunction as long as the one\hyp{}particle
density matrices are available. It is shown that it is possible to accurately predict 
the changes in the one\hyp{}electron ground\hyp{}state density of water molecule 
in spatially uniform electric field, as well as in spatially non\hyp{}uniform 
electric field distribution generated by point charges. When both linear 
and quadratic terms with respect to the electric field are accounted for,
the electric field\hyp{}induced polarization energies, dipole moments 
and quadrupole moments are quantitatively described by the present theory 
in electric fields from weak to very strong in magnitude (0.003 -- 0.07 a.u.).
It is believed that the proposed model could open new routes in Quantum Chemistry 
for fast and efficient calculations of molecular properties in condensed phases.
\end{abstract}

@@

The development of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM)
technique was a revolution in computational chemistry which allows insightful theoretical studies
of quantum mechanical effects in biological systems because the costly quantum mechanical calculations
could be combined with lower\hyp{}level and much less expensive
classical dynamics.\cite{Warshel.Levitt.JMolBiol.1976,Senn.Thiel.Angew.2009}
Understanding physics that rules intermolecular interactions\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} 
helped to develop
a wide family of force fields, and served as a basis for embedding Hamiltonian
within an effective environment and fragmentation methods.\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}

@@

Most of the efforts to theoretically treat extended molecular aggregates are rooted in 
energy\hyp{} (or force field\hyp{})based philosophy. However,
in the past decade there has been considerable interest in density\hyp{}based approaches
of a wider scope and aims.\cite{Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Mandado.Hermida-Ramon.JCTC.2011,
Sun.Chan.ACR.2016,
Hedegard.Reiher.JCTC.2016} 
In particular, the importance of
one\hyp{}electron densities in chemistry is well known and acknowledged in the community,\cite{Holas.March.PhysRevA.1991}
though obtaining the exact density has been an undersolved issue for most of chemical systems.
Given that fact, it would be incredibly insightful 
to establish a first\hyp{}principles connection between the one\hyp{}particle
density and the perturbation in the environment through introduction of
\emph{density matrix susceptibility} (DMS) tensors. 
The main postulate of the present work is that the associated change 
in the one\hyp{}particle density matrix, $\delta {\bf D}$, can be
represented in a product form of the generalized perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$,
evaluated at certain site $i$.
%
\begin{equation*} \label{e:initial-model}
  \delta {\bf D} \approx \sum_i^{\text{Sites}} {\bf B}_i
  \cdot  \delta {\bf P}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contains critical information 
about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed 
molecule and is to be precomputed only once. 
Thus, such DMS could be adapted into an efficient effective 
one\hyp{}electron fragment potential which in principle, is applicable to an arbitrary
level of theory, as long as the one\hyp{}particle densities are available.
It also means that such a technique would dramatically reduce costs of quantum chemistry calculations
because the task of performing full computation on a molecular aggregate
would reduce to much simpler matrix operations.

The simplest type of perturbation, yet the most relevant for chemistry, 
is the electrostatic perturbation.
It is worth mentioning that, at the moment, there is no simple \emph{ab initio}
method to compute electron deformation densities without considering the entire molecular 
aggregate by using full quantum chemistry calculations.\cite{Horn.Head-Gordon.JCP.2015,
Mandado.Hermida-Ramon.JCTC.2011}
In this work, we explore the susceptibility of the molecule's electronic density distribution
to polarize due to external electrostatic perturbation from first principles. 
The deformation density is expressed in terms
of the one\hyp{}particle density matrix polarization susceptibility tensors 
that are properties of isolated molecules and can be
extracted from quantum chemistry calculations. The focus on the density (matrix) instead of the energy
(scalar) is beneficial since it provides the directional tensor description
of the polarization that could be combined with other effects such as Pauli repulsion 
or dispersion,\cite{Mandado.Hermida-Ramon.JCTC.2011}
in contrast to the standard interaction energy approaches. We show that accurate polarization 
of the one\hyp{}particle density
matrix is achievable and can be realized as long as the one\hyp{}particle density matrices
are available. It is believed that the proposed method
could be a basis for a construction of an 
alternative approach in fragment\hyp{}based
modelling of extended molecular aggregates that is not limitted to the ground state chemistry.

@@

In the limit of a weak and spatially uniform electric field,
the polarization\hyp{}induced density matrix change is given by the following Taylor expansion:
%
\begin{equation}\label{e:dD-Taylor}
 \delta D_{\alpha\beta} = \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}}  \cdot {\bf F} 
   + \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} : {\bf F} \otimes {\bf F}
   + \ldots 
\end{equation}
%
In the above expression, the density matrix susceptibilities are simply equal to the derivatives of the density matrix
with respect to electric field (up to a multiplicative constant).
In a general case, i.e., when electric field is non\hyp{}uniform, the development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
becomes extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because
a working model has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element will accumulate when performing two\hyp{}
and four\hyp{}index contractions,
and in turn, affect the evaluated properties with non\hyp{}negligible errors.
Single\hyp{}centered expansions, like the one from Eq.~\eqref{e:dD-Taylor}, although valid for 
an arbitrary level of theory,
are not sufficient enough to accurately describe the local polarization
of electron density distribution when electric fields are highly non\hyp{}uniform in space.
Due to all of these reasons, we adopt the following approach: 
we first develop a qualitative first\hyp{}principles distributed\hyp{}site model at the Hartree\hyp{}Fock
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the density matrix susceptibility tensors as cetrain functions of adjustable parameters further on.

@@

$\delta {\bf D}$ should 
be directly related to the
polarization\hyp{}induced multipole moments because they are
a measure of the distortion of the electron density distribution due to the external electric
field. 
Therefore, 
we start our analysis by considering the induced electronic dipole moment 
of a closed-shell molecule, 
%
\begin{equation} \label{e:dmu-exact}
 \delta {\BM{\upmu}}({\bf r}_Q) = 
     2 \Tr{ 
         \left[ 
              \delta {\bf D} \cdot {\mathbb{M}({\bf r}_Q)}
         \right] } \;,
\end{equation}
%
where ${\mathbb{M}({\bf r}_Q)}$ is atomic orbital (AO)\hyp{}representation
of the dipole moment operator defined with respect to origin at ${\bf r}_Q$,
%
\begin{equation}\label{e:m}
 \left[ {\mathbb{M}({\bf r}_Q)} \right]_{\alpha\beta} = -e\tbraket{\alpha}{{\bf r} - {\bf r}_Q}{\beta} \;.
\end{equation}
%
(In this work we shall use the special open bold font for matrices of vectors, so that
in this case ${\mathbb{M}}$ can be considered as a $n \times n$ matrix of vectors of length 3).
Note that if the total charge is conserved, the induced moments are independent of the origin.

@@

Now, if the AO basis is orthogonal, Eq.~\eqref{e:dmu-4-exact.linear-approximation} can be simplified by realising that
%
\begin{equation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{M}} \cdot \delta {\bf C}
   \right] }  
-
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{K}} \cdot \delta {\bf C}
   \right] } \;.
\end{equation}
%
Here we introduced the auxiliary tensors
%
\begin{subequations}
 \begin{align}
   \widetilde{\mathbb{M}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}}     \;,           \\
   \widetilde{\mathbb{K}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}} \cdot {\bf D}^{(0)} \;, \\
   \widetilde{\mathbb{L}}  &\equiv \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \;
 \end{align}
\end{subequations}
%
to simplify the notation. Hence, we have
%
\begin{equation} \label{e:dmu-l-vector}
  \frac{1}{4} 
 \delta {\BM{\upmu}} %({\bf r}_0) 
   =
   \Tr{ 
    \left[ 
         \widetilde{\mathbb{L}} \cdot \delta {\bf C}
    \right] }
   %
   \equiv \sum_{i}^{\rm occ} \sum_\alpha {\bf l}_{i\alpha} \delta C_{\alpha i} \;,
\end{equation}
%
where the Cartesian vector
%
\begin{equation}\label{e:l-vector}
 {\bf l}_{i\alpha} \equiv \left[ \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \right]_{i\alpha} 
      = \left[  {\bf C}^\dagger \cdot \mathbb{M} \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \right]_{i\alpha}
\end{equation}
%
and $i$ runs over occupied unperturbed MO's.
Note that the intermediate quantity $\widetilde{\mathbb{L}}$ contains the projector onto
the unoccupied MO space at the absence of the external electric field. It means that the
perturbed solution is constructed from the unperturbed virtual orbitals as well.
Although Eq.~\eqref{e:dmu-l-vector}
is exact up to first order in ${\BM\Delta}$, it cannot be inverted at this moment to obtain $\delta {\bf C}$
due to the presence of trace operation, unless the electric field is uniform.
Let the induced dipole moment be partitioned into separate contributions
from the occupied MO's, i.e.,
%
\begin{equation} \label{e:dmu-part}
 \delta {\BM{\upmu}} = \sum_i^{\rm occ} \delta {\BM{\upmu}}_i \;,
\end{equation}
%
where
%
\begin{equation} \label{e:dmu-dpol}
 \delta {\BM{\upmu}}_i = {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) \;.
\end{equation}
%
In Eq.~\eqref{e:dmu-dpol}, ${\BM{\alpha}}_i$ is the distributed dipole\hyp{}dipole polarizability tensor
associated with the $i$th MO whereas ${\bf F}({\bf r}_i)$ is the electric field evaluated at the $i$th MO centroid of charge,
${\bf r}_i = \tbraket{\phi_i}{\hat{\bf r}}{\phi_i}$.
By inserting the above result into Eq.~\eqref{e:dmu-l-vector},
recasting the right hand side in a matrix form
and assuming real quantities,
one can separate the contributions coming from occupied MO's as
%
\begin{equation} \label{e:dmu-l-vector-mo-transform-X}
 \frac{1}{4} {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) 
   =
   \delta {\bf c}_i^T \cdot {\bf L}_i \;.
\end{equation}
%
Note that now ${\bf L}_i$ is a matrix of dimension $n \times 3$ and $\delta{\bf c}_i$ is the $i$th column of the
change in the LCAO-MO matrix due to the polarization process.
The solution for $\delta {\bf C}$ is
%
\begin{equation} \label{e:mu-ind.matrix.linear.solution}
  \delta {\bf c}_i^T = \frac{1}{4}
            {\bf F}({\bf r}_i)  \cdot {\BM{\alpha}}_i \cdot 
                    \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \;,
\end{equation}
%
where $\left[ {\bf L}_i  \right]^{-1}_{\rm Left}$ is a left inverse
of ${\bf L}_i$ matrix,
%
\begin{equation} 
      \left[ {\bf L}_i  \right]^{-1}_{\rm Left}   \equiv
       \left[ {\bf L}_i^T \cdot {\bf L}_i \right]^{-1} \cdot {\bf L}_i^T   
\end{equation}
%
with the dimension of $3\times n$. Note that the right inverse of ${\bf L}_i$
does not exist because ${\bf L}_i \cdot {\bf L}_i^T$ is singular due to the fact that usually
$n > 3$.

@@

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parameterized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha}^{(i;1)} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha;w}^{(i;1)}= \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. 
The superscript `$(1)$' underlines the fact that the susceptibility introduced here
is first\hyp{}order with respect to the electric field.
The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i;1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i;1)}_{\alpha\beta} = 
                               C_{\alpha i} {\bf b}_{\beta}^{(i;1)} + C_{\beta i} {\bf b}_\alpha^{(i;1)} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma}^{(i;1)}
 \;.
\end{multline}
%
Hence, the three dimensional vector ${\bf B}^{(i;1)}_{\alpha\beta}$ describes the polarization affinity
of the density matrix element associated with the $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all elements of ${\bf B}_{\alpha\beta}^{(i;1)}$
are unique properties of electronically unperturbed molecule.

One very important feature of the susceptibility from Eq.~\eqref{e:final-model.HF} is that
the density matrix increment is dependent \emph{only} on the susceptibility element associated with the same
pair of AO's. Therefore, there is no mixing in the AO's, rather the increment is proportional to the 
vector susceptibility, different for each density matrix element. This feature will be used to
generalize the model to an arbitrary level of theory and electric field curvature in the next section.

@@

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.\cite{Psi4.JCTC.2017}
Water molecule
was selected as a test system which was subjected to 
electrostatic perturbations due to random uniform electric fields and random collections of point charges
generating non\hyp{}uniform electric fields. 
Components of spatially uniform electric fields were determined 
from the uniform distribution in the interval of $\vert 0.0404 \vert$ a.u.
generating average field of 0.035 a.u. in an ensemble of 100 samples.
Cartesian coordinates of point charges were drawn from the uniform distribution 
of a set of fourty three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
with an origin that is located in the center of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Electric charge values were selected from the one\hyp{}dimensional uniform distribution
of double precision numbers in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly).
In this way, independent sets of 100 random samples of uniform electric fields as well as 
40 point charges were selected and exact solutions of HF\hyp{}SCF equations were 
performed for each validation procedure.

>>>>

The electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix 
susceptibilities (DMS) interacting with the spatial distribution of electric field.
The analytical approximate expressions are derived at the Hartree\hyp{}Fock theory, 
which serves as a basis for construction of generalized model 
that can be used for an arbitrary form of wavefunction and any type of one\hyp{}particle density matrix
as long as the latter are available by means of quantum chemistry methods. 
It is shown that it is possible to accurately predict 
the changes in the one\hyp{}electron ground\hyp{}state density of water molecule 
in spatially uniform electric field, as well as in spatially non\hyp{}uniform 
electric field distribution generated by point charges. When both linear 
and quadratic terms with respect to the electric field are accounted for,
the electric field\hyp{}induced polarization energies, dipole moments 
and quadrupole moments are quantitatively described by the present theory 
in electric fields ranging from weak to very strong (0.001 -- 0.07 a.u.).
It is believed that the proposed model could open new routes in quantum chemistry 
for fast and efficient calculations of molecular properties in condensed phases.
\end{abstract}

@@

The development of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM)
technique was a revolution in computational chemistry which allows insightful theoretical studies
of quantum mechanical effects in biological systems because the costly quantum mechanical calculations
can be combined with lower\hyp{}level and much less expensive
classical dynamics.\cite{Warshel.Levitt.JMolBiol.1976,Senn.Thiel.Angew.2009}
Understanding physics that rules intermolecular interactions\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} 
helped to develop
a wide family of force fields, and served as a basis for embedding Hamiltonian
within an effective environment and fragmentation methods.\cite{Gordon.Fedorov.Pruitt.Slipchenko.ChemRev.2012}

@@

Most of the efforts to theoretically treat extended molecular aggregates are rooted in 
energy\hyp{}based philosophy. However,
in the past decade there has been considerable interest in density\hyp{}based approaches
of a wider scope and aims.\cite{Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Mandado.Hermida-Ramon.JCTC.2011,
Sun.Chan.ACR.2016,
Hedegard.Reiher.JCTC.2016} 
In particular, the importance of
one\hyp{}electron densities in chemistry is well known and acknowledged in the community,\cite{Holas.March.PhysRevA.1991}
though obtaining the accurate electron density distribution has been an undersolved issue for most of chemical systems
in condensed phases.
Given that fact, it would be incredibly insightful 
to establish a first\hyp{}principles connection between the one\hyp{}particle
density and the perturbation in the environment through introduction of
\emph{density matrix susceptibility} (DMS) tensors. 
The main postulate of the present work is that the associated change 
in the one\hyp{}particle density matrix, $\delta {\bf D}$, can be
represented in a product form of the generalized perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$,
evaluated at certain site $i$.
%
\begin{equation*} \label{e:initial-model}
  \delta {\bf D} \approx \sum_i^{\text{Sites}} {\bf B}_i
  \cdot  \delta {\bf P}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contains critical information 
about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed 
molecule and is to be precomputed only once. 
Thus, such DMS could be adapted into an efficient effective 
one\hyp{}electron fragment potential which in principle, is applicable to an arbitrary
level of theory, as long as the one\hyp{}particle densities are available.
It also means that such a technique would dramatically reduce costs of quantum chemistry calculations
because the task of performing full computation on a molecular aggregate
would reduce to much simpler matrix operations.

The simplest type of perturbation, yet the most relevant for chemistry, 
is the electrostatic perturbation.
It is worth mentioning that, at the moment, there is no simple \emph{ab initio}
method to compute polarization\hyp{}induced electron deformation densities 
without considering the entire molecular 
aggregate by using full quantum chemistry calculations.\cite{Horn.Head-Gordon.JCP.2015,
Mandado.Hermida-Ramon.JCTC.2011}
In this work, we explore the susceptibility of the molecule's electronic density distribution
to polarize due to external electrostatic perturbation from first principles. 
The deformation density is expressed in terms
of the one\hyp{}particle density matrix polarization susceptibility tensors 
that are properties of isolated molecules and can be
extracted from quantum chemistry calculations. The focus on the density (matrix) instead of the energy
(scalar) is beneficial since it provides the directional tensor description
of the polarization that could be combined with other effects such as Pauli repulsion 
or dispersion,\cite{Mandado.Hermida-Ramon.JCTC.2011}
in contrast to the standard interaction energy approaches. We show that accurate polarization 
of the one\hyp{}particle density
matrix is achievable and can be realized as long as the one\hyp{}particle density matrices
are available. It is believed that the proposed method
could be a basis for a construction of an 
alternative approach in fragment\hyp{}based
modelling of extended molecular aggregates that is not limitted to the ground state chemistry.

@@

In the limit of a weak and spatially uniform electric field,
the polarization\hyp{}induced density matrix change of any electronic state
or electronic transition is given by the following Taylor expansion:
%
\begin{equation}\label{e:dD-Taylor}
 \delta D_{\alpha\beta} = \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}}  \cdot {\bf F} 
   + \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} : {\bf F} \otimes {\bf F}
   + \ldots 
\end{equation}
%
In the above expression, the density matrix susceptibilities 
are simply equal to the derivatives of the density matrix
with respect to electric field (up to a multiplicative constant).
In a general case, i.e., when electric field is relatively strong 
and non\hyp{}uniform, single\hyp{}centered expansions, like the one 
from Eq.~\eqref{e:dD-Taylor},
%although valid for an arbitrary level of theory in uniform fields,
are not sufficient enough to accurately describe the local polarization
of electron density distribution.

The development of a general first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence 
of the one\hyp{}particle density becomes extremely difficult. This is not 
only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because a working model 
has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element will accumulate when performing two\hyp{}
and four\hyp{}index contractions, and in turn, affect the evaluated properties 
with non\hyp{}negligible errors.

Due to all of these reasons, we adopt the following approach: 
we first develop a qualitative \emph{ab initio} distributed\hyp{}site 
model at the Hartree\hyp{}Fock level of theory. Further on, we generalize it 
to an arbitrary level and type of density matrix
using the functional form that was found by considering HF level as a guideline,
but treating the density matrix susceptibility tensors as cetrain functions 
of adjustable parameters.

@@

$\delta {\bf D}$ from Eq.~\eqref{e:rho-1} should 
be directly related to the
polarization\hyp{}induced multipole moments because they are
a measure of the distortion of the electron density distribution due to the external electric
field. 
Therefore, we start our analysis by considering the induced 
ground\hyp{}state electronic dipole moment of a closed\hyp{}shell molecule, 
%
\begin{equation} \label{e:dmu-exact}
 \delta {\BM{\upmu}}({\bf r}_Q) = 
     2 \Tr{ 
         \left[ 
              \delta {\bf D} \cdot {\mathbb{M}({\bf r}_Q)}
         \right] } \;,
\end{equation}
%
where ${\mathbb{M}({\bf r}_Q)}$ is atomic orbital (AO)\hyp{}representation
of the dipole moment operator defined with respect to origin at ${\bf r}_Q$,
%
\begin{equation}\label{e:m}
 \left[ {\mathbb{M}({\bf r}_Q)} \right]_{\alpha\beta} = -\tbraket{\alpha}{{\bf r} - {\bf r}_Q}{\beta} \;.
\end{equation}
%
(In this work we shall use the double struck font for matrices of vectors, so that
in this case ${\mathbb{M}}$ can be considered as a $n \times n$ matrix of vectors of length 3).
Note that if the total charge is conserved, the induced moments are independent of the origin.

@@

Now, if the AO basis is orthogonal, Eq.~\eqref{e:dmu-4-exact.linear-approximation} can be simplified by realising that
%
\begin{equation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{M}} \cdot \delta {\bf C}
   \right] }  
-
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{K}} \cdot \delta {\bf C}
   \right] } \;.
\end{equation}
%
Here we introduced the auxiliary tensors
%
\begin{subequations}
 \begin{align}
   \widetilde{\mathbb{M}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}}     \;,           \\
   \widetilde{\mathbb{K}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}} \cdot {\bf D}^{(0)} \;, \\
   \widetilde{\mathbb{L}}  &\equiv \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \;
 \end{align}
\end{subequations}
%
to simplify the notation. Hence, we have
%
\begin{equation} \label{e:dmu-l-vector}
  \frac{1}{4} 
 \delta {\BM{\upmu}} %({\bf r}_0) 
   =
   \Tr{ 
    \left[ 
         \widetilde{\mathbb{L}} \cdot \delta {\bf C}
    \right] }
   %
   \equiv \sum_{i}^{\rm occ} \sum_\alpha {\bf l}_{i\alpha} \delta C_{\alpha i} \;,
\end{equation}
%
where the Cartesian vector
%
\begin{equation}\label{e:l-vector}
 {\bf l}_{i\alpha} \equiv \left[ \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \right]_{i\alpha} 
      = \left[  {\bf C}^\dagger \cdot \mathbb{M} \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \right]_{i\alpha}
\end{equation}
%
and $i$ runs over occupied unperturbed MO's.
Note that the intermediate quantity $\widetilde{\mathbb{L}}$ contains the projector onto
the unoccupied MO space at the absence of the external electric field. It means that the
perturbed solution is constructed from the unperturbed virtual orbitals as well.
Although Eq.~\eqref{e:dmu-l-vector}
is exact up to first order in ${\BM\Delta}$, it cannot be inverted at this moment to obtain $\delta {\bf C}$
due to the presence of trace operation, unless the electric field is uniform.
Let the induced dipole moment be partitioned into separate contributions
from the occupied localized MO's (LMO's), i.e.,
%
\begin{equation} \label{e:dmu-part}
 \delta {\BM{\upmu}} = \sum_i^{\rm occ} \delta {\BM{\upmu}}_i \;,
\end{equation}
%
where
%
\begin{equation} \label{e:dmu-dpol}
 \delta {\BM{\upmu}}_i = {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) \;.
\end{equation}
%
In Eq.~\eqref{e:dmu-dpol}, ${\BM{\alpha}}_i$ is the distributed dipole\hyp{}dipole polarizability tensor
associated with the $i$th LMO whereas ${\bf F}({\bf r}_i)$ 
is the electric field evaluated at the $i$th LMO charge centroid,
${\bf r}_i = \tbraket{\phi_i}{\hat{\bf r}}{\phi_i}$.
By inserting the above result into Eq.~\eqref{e:dmu-l-vector},
recasting the right hand side in a matrix form
and assuming real quantities,
one can separate the contributions coming from occupied MO's as
%
\begin{equation} \label{e:dmu-l-vector-mo-transform-X}
 \frac{1}{4} {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) 
   =
   \delta {\bf c}_i^T \cdot {\bf L}_i \;.
\end{equation}
%
Note that now ${\bf L}_i$ is a matrix of dimension $n \times 3$ and $\delta{\bf c}_i$ is the $i$th column of the
change in the LCAO-MO matrix due to the polarization process.
The solution for $\delta {\bf C}$ is
%
\begin{equation} \label{e:mu-ind.matrix.linear.solution}
  \delta {\bf c}_i^T = \frac{1}{4}
            {\bf F}({\bf r}_i)  \cdot {\BM{\alpha}}_i \cdot 
                    \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \;,
\end{equation}
%
where $\left[ {\bf L}_i  \right]^{-1}_{\rm Left}$ is a left inverse
of ${\bf L}_i$ matrix,
%
\begin{equation} 
      \left[ {\bf L}_i  \right]^{-1}_{\rm Left}   \equiv
       \left[ {\bf L}_i^T \cdot {\bf L}_i \right]^{-1} \cdot {\bf L}_i^T   
\end{equation}
%
with the dimension of $3\times n$. Note that the right inverse of ${\bf L}_i$
does not exist because ${\bf L}_i \cdot {\bf L}_i^T$ is singular due to the fact that usually
$n > 3$.

@@

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parameterized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha}^{(i;1)} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha;w}^{(i;1)}= \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. 
The superscript `$(1)$' underlines the fact that the susceptibility introduced here
is first\hyp{}order with respect to the electric field.
The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i;1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i;1)}_{\alpha\beta} = 
                               C_{\alpha i} {\bf b}_{\beta}^{(i;1)} + C_{\beta i} {\bf b}_\alpha^{(i;1)} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma}^{(i;1)}
 \;.
\end{multline}
%
Hence, the three dimensional vector ${\bf B}^{(i;1)}_{\alpha\beta}$ 
describes the local linear polarization affinity 
of the density matrix element 
associated with the $i$th LMO centroid as well as the $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all the elements of ${\bf B}_{\alpha\beta}^{(i;1)}$
are unique properties of electronically unperturbed molecule. 
However, they are valid only under the Hartree\hyp{}Fock
approximation, with a guarantee that only the overall charge (exactly) and the dipole moment (approximately,
limited by the flavour of the distributed polarizability approximation), 
but not higher multipole moments and polarization energy, are correctly described.

Nevertheless, one very important feature of the susceptibility from Eq.~\eqref{e:final-model.HF} is that
the density matrix increment is dependent \emph{only} on the susceptibility element associated with the same
pair of AO's. Therefore, there is no mixing in the AO's, rather the increment is proportional to the 
vector susceptibility, different for each density matrix element. This feature will be used to
generalize the model to an arbitrary level of theory and order of electric field response in the next section.

@@

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.\cite{Psi4.JCTC.2017}
Water molecule was selected as a test system which was subjected to 
electrostatic perturbations due to random uniform electric fields and random collections of point charges
generating non\hyp{}uniform electric fields.
Components of spatially uniform electric fields were determined 
from the uniform distribution in the interval 
of $\vert 0.0404 \vert$ a.u. generating average field of 0.035 a.u. in an ensemble of 100 samples.
Cartesian coordinates of point charges were drawn from the uniform distribution 
of a set of fourty three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
with an origin that is located in the center of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Electric charge values were selected from the one\hyp{}dimensional uniform distribution
in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly).
In this way, independent sets of 100 random samples of either uniform electric field or
point charge distributions were selected and exact solutions of HF\hyp{}SCF equations were 
performed for each validation procedure.
In the case of non\hyp{}uniform electric fields, the distribution centres 
of the \emph{ab initio} DMS model (Eq.~\eqref{e:final-model.HF}) were assumed 
to be LMO centroids, while atoms were used for the generalized DMS models (Eq.~\eqref{e:final-model.General}).
This is because various LMO centroids are often located at the same point which would
introduce linear dependencies in the Hessian matrix from Eq.~\eqref{e:Newton.gH}.
